Dynamical properties of the Benguela Current

### 2019-05-09 Paper1_results.Rmd
exploring percentile bits

library(tidyverse)
library(plyr)
library(lubridate)
library(ggpubr)
library(zoo)
library(lubridate)
library(FNN)
library(forecast)
library(astrochron)
library(WaveletComp)
library(data.table)
library(heatwaveR)
library(viridis)
library(ggrepel)
library(plyr)
library(maptools)
library(sp)
library(geosphere)
library(PBSmapping)
library(scales)
library(grid)
library(gridExtra)
library(circular)
library(fossil)
library(mapproj)
library(ncdf4)
library(reshape2)
library(plyr) # Never load plyr when also loading the tidyverse. It causes a lot of conflicts.
library(stringr)
library(doMC); doMC::registerDoMC(cores = 4)
library(fasttime)
library(xtable)
library(ggpubr)

# library(devtools)
# install_github("marchtaylor/sinkr")

@@

# png("Figures/site_map.png", width = 8, height = 5, pointsize = 6)
```

@@

daily_chloro <- chloro_data %>%
  # dplyr::mutate(date = as.Date(date)) %>% 
  dplyr::group_by(combi, lon, lat, date) %>%
  dplyr::summarise(chloro = mean(chloro, na.rm = TRUE)) %>%
  dplyr::group_by(combi, lon, lat) %>% 
  drop_na() %>% 
  tidyr::nest() %>%
  dplyr::mutate(clims = purrr::map(data, ts2clm, x = date, y = chloro,
                     climatologyPeriod = c("2010-02-11", "2014-12-20"),
                     maxPadLength = 8)) %>% # Adjust maxPadLength = X for longer interpolation
  dplyr::select(-data) %>%
  tidyr::unnest() %>%
  select(-doy, -seas, -thresh)
```

@@

# Here I want to test using generalised linear hypotheses if there is a difference in the intensity of upwellling events between the insitu and       satellite data.
# Do the in situ and satelllite derived data yield the same number of events. First I add a seasonal column to test seasonal variation between 
library(multcomp) # for generalized linear hypothesis test (glht)

```

Determining the R2 value between *in situ* and remotely sensed SST

```{r}
# Loading the *in situ* and Remotely sensed SST

load("Data_P1/SACTN_fyears.RData")
load("Data_P1/OISST_fyears.RData")

R2 <- data.frame()
for(i in 1:length(levels(SACTN_fyears$site))) {
  x <- subset(SACTN_fyears, site == levels(SACTN_fyears$site)[i])
  y <- subset(df, site == levels(df$site)[i])
  x <- x[x$date %in% y$date,]
  y <- y[y$date %in% x$date,]
  z <- data.frame(site = as.character(x$site[1]), R2 = coef(lm(y$temp~x$temp))[2])
  R2 <- rbind(R2, z)
}
```

# Upwelling index
## Determining upwelling index from wind data (WindGuru and SAWS)
## Index Equation from Fielding & Davis 1989 paper

@@

# Seperating sites
index_SP <- index %>% 
  filter(site == "Sea Point") %>% 
  slice(-(c(34:70,435:632,686,1704)))

@@

index_plot_SP <- ggplot(UIi_SP, aes(x=Day, y=Index)) +
  geom_area(aes(fill=saws_condition)) +
  geom_hline(yintercept=0) +
  theme_bw() +
  # annotate("text", x = 300, y = -12.5, label = "SAWS Data") +
  #geom_vline(xintercept = c(116, 117, 133, 134, 244, 291), lty = 2, lwd = 1) +
  scale_fill_manual(values=c("salmon", "steelblue")) +
  scale_x_continuous(expand=c(0, 0)) +
  scale_y_reverse() +
  labs(y= "Upwelling Index") +
  guides(fill=guide_legend(title="Condition")) +
  ggtitle("Sea Point")
index_plot_SP

@@

index_plot_KOM <- ggplot(UIi_KOM, aes(x=Day, y=Index)) +
  geom_area(aes(fill=saws_condition)) +
  geom_hline(yintercept=0) +
  theme_bw() +
  # annotate("text", x = 300, y = -12.5, label = "SAWS Data") +
  #geom_vline(xintercept = c(116, 117, 133, 134, 244, 291), lty = 2, lwd = 1) +
  scale_fill_manual(values=c("salmon", "steelblue")) +
  scale_x_continuous(expand=c(0, 0)) +
  scale_y_reverse() +
  labs(y= "Upwelling Index") +
  guides(fill=guide_legend(title="Condition")) +
  ggtitle("Kommetjie")
index_plot_KOM

@@

index_plot_PN<- ggplot(UIi_PN, aes(x=Day, y=Index)) +
  geom_area(aes(fill=saws_condition)) +
  geom_hline(yintercept=0) +
  theme_bw() +
  # annotate("text", x = 300, y = -12.5, label = "SAWS Data") +
  #geom_vline(xintercept = c(116, 117, 133, 134, 244, 291), lty = 2, lwd = 1) +
  scale_fill_manual(values=c("salmon", "steelblue")) +
  scale_x_continuous(expand=c(0, 0)) +
  scale_y_reverse() +
  labs(y= "Upwelling Index") +
  guides(fill=guide_legend(title="Condition")) +
  ggtitle("Port Nolloth")
index_plot_PN

@@

index_plot_YZ<- ggplot(UIi_YZ, aes(x=Day, y=Index)) +
  geom_area(aes(fill=saws_condition)) +
  geom_hline(yintercept=0) +
  theme_bw() +
  # annotate("text", x = 300, y = -12.5, label = "SAWS Data") +
  #geom_vline(xintercept = c(116, 117, 133, 134, 244, 291), lty = 2, lwd = 1) +
  scale_fill_manual(values=c("salmon", "steelblue")) +
  scale_x_continuous(expand=c(0, 0)) +
  scale_y_reverse() +
  labs(y= "Upwelling Index") +
  guides(fill=guide_legend(title="Condition")) +
  ggtitle("Yzerfontein")
index_plot_YZ

@@

index_plot_LB<- ggplot(UIi_LB, aes(x=Day, y=Index)) +
  geom_area(aes(fill=saws_condition)) +
  geom_hline(yintercept=0) +
  theme_bw() +
  # annotate("text", x = 300, y = -12.5, label = "SAWS Data") +
  #geom_vline(xintercept = c(116, 117, 133, 134, 244, 291), lty = 2, lwd = 1) +
  scale_fill_manual(values=c("salmon", "steelblue")) +
  scale_x_continuous(expand=c(0, 0)) +
  scale_y_reverse() +
  labs(y= "Upwelling Index") +
  guides(fill=guide_legend(title="Condition")) +
  ggtitle("Lamberts Bay")
index_plot_LB

@@

index_plot_SB<- ggplot(UIi_SB, aes(x=Day, y=Index)) +
  geom_area(aes(fill=saws_condition)) +
  geom_hline(yintercept=0) +
  theme_bw() +
  # annotate("text", x = 300, y = -12.5, label = "SAWS Data") +
  #geom_vline(xintercept = c(116, 117, 133, 134, 244, 291), lty = 2, lwd = 1) +
  scale_fill_manual(values=c("salmon", "steelblue")) +
  scale_x_continuous(expand=c(0, 0)) +
  scale_y_reverse() +
  labs(y= "Upwelling Index") +
  guides(fill=guide_legend(title="Condition")) +
  ggtitle("Saldanha Bay")
index_plot_SB

upwelling_index <- ggarrange(index_plot_SP, index_plot_KOM, index_plot_PN,
          index_plot_YZ,index_plot_LB, index_plot_SB, ncol = 2, nrow = 3)
```

Calculating the temperature anomalies. This will be plotted on the same graph as the upwelling indeces. This allows for an observation on the relationship or patterns that exist as a result of temperature changes and he upwelling and downwelling conditions.

```

### EOF's

```{r} 
PCA_1 <- rda(tester)
summary(PCA)

# Proportion of variation explained
# Values to be found opon running the code

PCA <- cbind(scores(PCA_1, choices = c(1:4), display = "sites"))
PCA_spe <- cbind(scores(PCA_1, choices = c(1:4), display = "species"))

test1 <- ggplot(PCA, aes(x = lon, y = lat)) +
  geom_raster(aes(fill = PC1)) +
  geom_polygon(data = PC1, aes(long, lat, group = group),
               fill = NA, colour = "blue", size = 0.3) +
  geom_polygon(data = velocity, aes(long, lat, group = group),
               fill = NA, colour = "green", size = 0.3)+ 
         fill = guide_colourbar(title = "PC1")

test2 <- ggplot(PCA, aes(x = lon, y = lat)) +
  geom_raster(aes(fill = PC2)) +
  geom_polygon(data = PC1, aes(long, lat, group = group),
               fill = NA, colour = "blue", size = 0.3) +
  geom_polygon(data = velocity, aes(long, lat, group = group),
               fill = NA, colour = "green", size = 0.3)+ 
         fill = guide_colourbar(title = "PC2")

test3 <- ggplot(PCA, aes(x = lon, y = lat)) +
  geom_raster(aes(fill = PC2)) +
  geom_polygon(data = PC1, aes(long, lat, group = group),
               fill = NA, colour = "blue", size = 0.3) +
  geom_polygon(data = velocity, aes(long, lat, group = group),
               fill = NA, colour = "green", size = 0.3)+ 
         fill = guide_colourbar(title = "PC3")

test4 <- ggplot(PCA, aes(x = lon, y = lat)) +
  geom_raster(aes(fill = PC1)) +
  geom_polygon(data = PC1, aes(long, lat, group = group),
               fill = NA, colour = "blue", size = 0.3) +
  geom_polygon(data = velocity, aes(long, lat, group = group),
               fill = NA, colour = "green", size = 0.3)+ 
         fill = guide_colourbar(title = "PC4")

pc1_combined <- ggarrange( test1, test2, test3, test4)

```

In particular, trends will be observed to be different at ocean and coastal locations and we hypothesize that those differences are due to changes in upwelling intensity.

>>>>

library(tidyverse)
library(plyr)
library(lubridate)
library(ggpubr)
library(zoo)
library(lubridate)
library(FNN)
library(forecast)
library(astrochron)
library(WaveletComp)
library(data.table)
library(heatwaveR)
library(viridis)
library(ggrepel)
library(plyr)
library(maptools)
library(sp)
library(geosphere)
library(PBSmapping)
library(scales)
library(grid)
library(gridExtra)
library(circular)
library(fossil)
library(mapproj)
library(ncdf4)
library(reshape2)
library(plyr) # Never load plyr when also loading the tidyverse. It causes a lot of conflicts.
library(stringr)
library(doMC); doMC::registerDoMC(cores = 4)
library(fasttime)
library(xtable)
library(ggpubr)
library(ggpubr); theme_set(theme_pubr())# library(devtools)
# install_github("marchtaylor/sinkr")

@@

# Site map
# The insetting needs to be placed at appropriate location
```

@@

daily_chloro <- chloro_data %>%
  # dplyr::mutate(date = as.Date(date)) %>% 
  dplyr::group_by(combi, lon, lat, date) %>%
  dplyr::summarise(chloro = mean(chloro, na.rm = TRUE)) %>%
  dplyr::group_by(combi, lon, lat) %>% 
  drop_na() %>% 
  tidyr::nest() %>%
  dplyr::mutate(clims = purrr::map(data, ts2clm, x = date, y = chloro,
                     climatologyPeriod = c("2010-02-11", "2014-12-20"),
                     maxPadLength = 8)) %>% # Adjust maxPadLength = X for longer interpolation
  dplyr::select(-data) %>%
  tidyr::unnest() %>%
  select(-doy, -seas, -thresh)

****Error: Evaluation error: sort(): detected NaN.*** 
```

@@

# Upwelling index
## Determining upwelling index from wind data (SAWS)
## Index Equation from Fielding & Davis 1989 paper
## This is only for the SACTN data

@@

# Seperating sites
index_SP <- index %>% 
  filter(site == "Sea Point") %>% 
  ungroup()#%>% 
  # slice(-(c(34:70,435:632,686,1704)))

@@

index_plot_SP <- ggplot(UIi_SP, aes(x=Day, y=Index)) +
  geom_area(aes(fill=saws_condition)) +
  geom_hline(yintercept=0) +
  theme_bw() +
  scale_fill_manual(values=c("salmon", "steelblue")) +
  scale_x_continuous(expand=c(0, 0)) +
  scale_y_reverse() +
  labs(y= "Upwelling Index") +
  guides(fill=guide_legend(title="Condition")) +
  ggtitle("Sea Point")
index_plot_SP

@@

index_plot_KOM <- ggplot(UIi_KOM, aes(x=Day, y=Index)) +
  geom_area(aes(fill=saws_condition)) +
  geom_hline(yintercept=0) +
  theme_bw() +
  scale_fill_manual(values=c("salmon", "steelblue")) +
  scale_x_continuous(expand=c(0, 0)) +
  scale_y_reverse() +
  labs(y= "Upwelling Index") +
  guides(fill=guide_legend(title="Condition")) +
  ggtitle("Kommetjie")
index_plot_KOM

@@

index_plot_PN<- ggplot(UIi_PN, aes(x=Day, y=Index)) +
  geom_area(aes(fill=saws_condition)) +
  geom_hline(yintercept=0) +
  theme_bw() +
  scale_fill_manual(values=c("salmon", "steelblue")) +
  scale_x_continuous(expand=c(0, 0)) +
  scale_y_reverse() +
  labs(y= "Upwelling Index") +
  guides(fill=guide_legend(title="Condition")) +
  ggtitle("Port Nolloth")
index_plot_PN

@@

index_plot_YZ<- ggplot(UIi_YZ, aes(x=Day, y=Index)) +
  geom_area(aes(fill=saws_condition)) +
  geom_hline(yintercept=0) +
  theme_bw() +
  scale_fill_manual(values=c("salmon", "steelblue")) +
  scale_x_continuous(expand=c(0, 0)) +
  scale_y_reverse() +
  labs(y= "Upwelling Index") +
  guides(fill=guide_legend(title="Condition")) +
  ggtitle("Yzerfontein")
index_plot_YZ

@@

index_plot_LB<- ggplot(UIi_LB, aes(x=Day, y=Index)) +
  geom_area(aes(fill=saws_condition)) +
  geom_hline(yintercept=0) +
  theme_bw() +
  scale_fill_manual(values=c("salmon", "steelblue")) +
  scale_x_continuous(expand=c(0, 0)) +
  scale_y_reverse() +
  labs(y= "Upwelling Index") +
  guides(fill=guide_legend(title="Condition")) +
  ggtitle("Lamberts Bay")
index_plot_LB

@@

index_plot_SB<- ggplot(UIi_SB, aes(x=Day, y=Index)) +
  geom_area(aes(fill=saws_condition)) +
  geom_hline(yintercept=0) +
  theme_bw() +
  scale_fill_manual(values=c("salmon", "steelblue")) +
  scale_x_continuous(expand=c(0, 0)) +
  scale_y_reverse() +
  labs(y= "Upwelling Index") +
  guides(fill=guide_legend(title="Condition")) +
  ggtitle("Saldanha Bay")
index_plot_SB

upwelling_index <- ggarrange(index_plot_SP, index_plot_KOM, index_plot_PN,
          index_plot_YZ,index_plot_LB, index_plot_SB, ncol = 2, nrow = 3)

# See graph: ("Figures/upwelling_indeces.png")
# Sea Point appears to have missing data
# Lamberts bay appears to have missing data as well
# Can we plot the graph where it ignores the missing data without interpolations? [AJ suggested]
# Plot the temperature anomalies on the same graphs as the upwelling indeces for each of the sites
```
######################################

# Question: How does the intensity and duration of an upwelling event vary at different distances from the coastline???
  # Got the various pixels : load(Data_P1/site_pixels.RData)
  # How do i determine where upwelling occured? - What is the threshold of an upwelling event? # How do i determine the threshold of this event
  # Calculate event count, duration and mean intensity for each pixel

Look at the percentile level that seawater temperature must decrease below when an upwelling event occurs
  - To what exttent the temperature of upwelled water  is colder than the mean SST. 
  - How the minumum temperature of upwelled water becomes to be that temperature
  - Why temperature does not become any colder

To do:
- how count, duration, intensity of upwelling changed between 1990s and current times

### 2019-05-09 Write_up/Writing/dissertation.Rmd
exploring percentile bits

Significant research studies accept that climate change and variability have great impacts on marine ecosystems [@stenseth; @harley; @hoegh]. The Intergovernmental Panel on Climate Change (IPCC) predicts future projections with a mean increase in air temperature of 1.1°C by the 2020’s and 1.8°C by the 2040’s as a result of the increase of greenhouse gas concentrations [@solomon]. A large amount of this heat will be absorbed by the world’s ocean. In current years, as a result of changes in air temperature, the heat content of the ocean has increased significantly. Given ocean warming in association with their physical changes such as sea level rise and declines in sea ice, marine life is expected to intensify in future decades. Global warming represents distinct warming and cooling periods throughout the 20th and 21st century (IPCC, 2007). Global warming however is not evenly distributed across the global ocean, and thus several regions are warming faster or slower than García-Reyes et al., 2015 the global average [@levitus; @santos]. In particular, the Atlantic Ocean contributes most to the increase of heat content [@nerem; @levitus].

@@

Climate change has severe ecological consequences on both plants and animals (Bradley et al. 1999; Walther et al. 2002). In the past two decades, sea surface temperature has risen in many regions and this Seasonal rise in temperature can act as an important cue in species survival, reproduction and abundance (Cushing 1982; Cushing 1990). The different responses of species as a result of climate change are important to understand as these changes affect, the trophic dynamics and fisheries yield of several regional ecosystems among many things (Murawski 1993). Understanding the relationship between global warming and climate change as a result of anthropogenic activity is a vast and challenging process. In a bid to further this understanding, Rockström et al. (2009) introduced the concept of planetary boundaries. According to this concept, within each of these boundaries exists a level of safety that should not be exceeded. According to this concept, the planetary boundary for biodiversity loss, has been surpassed [@steffen]. Similarly, the safety boundary for warming of the globe by greenhouse gas emission has also been surpassed. The idea of planetary boundaries is essential in that it helps in understanding the threats humans cause via the continued destruction of the natural environment. Consequently, studies of anthropogenic influences on the ecosystem are generally divided into land or sea. Terrestrial ecosystems are more familiar to the general populace; whereas marine ecosystems are associated with high levels of primary production [@Chavez] and are highly at risk as a result of climate change [@osterblom]. Climate change has been proven to decrease primary production thus affecting trophic levels and leading to trophic turbulence and the destruction of marine fauna [@lewandowska; @dell, @mcCauley].

@@

Northward, the Benguela Current Upwelling system is linked by the Angola Benguela Front which converge between the Benguela current and the Benguela coastal current as well as the warm saline waters of the Angola Current [@peterson]. Research done by Yamagata and Iizuka (1995) discovered that coastal Kelvin waves, originating from equatorial Kelvin waves, carry the down welling process southward.  Southward, the coastal upwelling system is bounded by the Agulhas Bank. South of the African coast the warm Agulhas Current, western boundary current, flows in from the Indian Ocean and returns eastwards while generating warm eddies that spin off in a north-westerly direction in the Benguela current [@duncombe]. The region surrounding the Benguela experiences a persistent along shore wind. This wind is associated with the St. Helena high pressure system. Lutjeharms and Meeuwis, (1987) also found a strong relationship between the intensity of upwelling events and the strength and direction of prevailing winds. These favourable upwelling equatorial winds represent a maximum at 25°S and decrease towards the northern and southern boundaries of the Benguela upwelling system. Continental shelf bathymetry and upwelling favourable winds provide a large-scale upwelling mechanism in the southern Benguela, whereas local topography and meteorology create an alternating pattern of active and passive upwelling circulations along the coast. Winds occurring within the southern regions represent a seasonal variation, reaching a maximum during spring and summer months [@boyd; @shannon1996; @fennel]. Winds occurring within this region tend to show increasing patterns away from the coast whereas, winds occurring in the northern region exhibits relatively less seasonal variation [@shannon1985; @pitcher]. Upwelling occurring along the south coast of the Cape Province within South Africa, specifically along the coastline is generated by local wind, with wind increasing somewhat away from the coast. This facilitated the idea of a wind stress curl, as discussed in Bakun and Nelson (1991) and Shannon and Nelson (1996).

>>>>

Significant research studies accept that climate change and variability have great impacts on marine ecosystems [@stenseth; @harley; @hoegh]. The Intergovernmental Panel on Climate Change (IPCC) predicts future projections with a mean increase in air temperature of 1.1°C by the 2020’s and 1.8°C by the 2040’s as a result of the increase of greenhouse gas concentrations [@solomon]. A large amount of this heat will be absorbed by the world’s ocean. In current years, as a result of changes in air temperature, the heat content of the ocean has increased significantly. Given ocean warming in association with their physical changes such as sea level rise and declines in sea ice, marine life is expected to intensify in future decades. Global warming represents distinct warming and cooling periods throughout the 20th and 21st century (IPCC, 2007). Global warming however is not evenly distributed across the global ocean, and thus several regions are warming faster or slower than the global average [@levitus; @santos]. In particular, the Atlantic Ocean contributes most to the increase of heat content [@nerem; @levitus].

@@

Climate change has severe ecological consequences on both plants and animals (Bradley et al. 1999; Walther et al. 2002). In the past two decades, sea surface temperature has risen in many regions and this seasonal rise in temperature can act as an important cue in species survival, reproduction and abundance (Cushing 1982; Cushing 1990). The different responses of species as a result of climate change are important to understand as these changes affect, the trophic dynamics and fisheries yield of several regional ecosystems among many things (Murawski 1993). Understanding the relationship between global warming and climate change as a result of anthropogenic activity is a vast and challenging process. In a bid to further this understanding, Rockström et al. (2009) introduced the concept of planetary boundaries. According to this concept, within each of these boundaries exists a level of safety that should not be exceeded. According to this concept, the planetary boundary for biodiversity loss, has been surpassed [@steffen]. Similarly, the safety boundary for warming of the globe by greenhouse gas emission has also been surpassed. The idea of planetary boundaries is essential in that it helps in understanding the threats humans cause via the continued destruction of the natural environment. Consequently, studies of anthropogenic influences on the ecosystem are generally divided into land or sea. Terrestrial ecosystems are more familiar to the general populace; whereas marine ecosystems are associated with high levels of primary production [@Chavez] and are highly at risk as a result of climate change [@osterblom]. Climate change has been proven to decrease primary production thus affecting trophic levels and leading to trophic turbulence and the destruction of marine fauna [@lewandowska; @dell, @mcCauley].

@@

Northward, the Benguela Current Upwelling system is linked by the Angola Benguela Front which converge between the Benguela current and the Benguela coastal current as well as the warm saline waters of the Angola Current [@peterson]. Research done by Yamagata and Iizuka (1995) discovered that coastal Kelvin waves, originating from equatorial Kelvin waves, carry the downwelling process southward.  Southward, the coastal upwelling system is bounded by the Agulhas Bank. South of the African coast the warm Agulhas Current, western boundary current, flows in from the Indian Ocean and returns eastwards while generating warm eddies that spin off in a north-westerly direction in the Benguela current [@duncombe]. The region surrounding the Benguela experiences a persistent along shore wind. This wind is associated with the St. Helena high pressure system. Lutjeharms and Meeuwis, (1987) also found a strong relationship between the intensity of upwelling events and the strength and direction of prevailing winds. These favourable upwelling equatorial winds represent a maximum at 25°S and decrease towards the northern and southern boundaries of the Benguela upwelling system. Continental shelf bathymetry and upwelling favourable winds provide a large-scale upwelling mechanism in the southern Benguela, whereas local topography and meteorology create an alternating pattern of active and passive upwelling circulations along the coast. Winds occurring within the southern regions represent a seasonal variation, reaching a maximum during spring and summer months [@boyd; @shannon1996; @fennel]. Winds occurring within this region tend to show increasing patterns away from the coast whereas, winds occurring in the northern region exhibits relatively less seasonal variation [@shannon1985; @pitcher]. Upwelling occurring along the south coast of the Cape Province within South Africa, specifically along the coastline is generated by local wind, with wind increasing somewhat away from the coast. This facilitated the idea of a wind stress curl, as discussed in Bakun and Nelson (1991) and Shannon and Nelson (1996).
